---
title: RMBL Climate Graph
author: Ian Breckheimer
date: today
format: html
---

```{r setup}

library(ggplot2)
library(dplyr)

setwd("~/code/demo_graphing_project/")
```

# Question

How has the hydroclimate of the RMBL vicinity changed over the past 8 years?

# Data

The data come from the GridMET dataset (Abatzaglou et al. 2013, https://www.researchgate.net/profile/John-Abatzoglou/publication/236687227_Development_of_gridded_surface_meteorological_data_for_ecological_applications_and_modeling/links/5a563217aca272bb6963c2ce/Development-of-gridded-surface-meteorological-data-for-ecological-applications-and-modeling.pdf).

I extracted values from the 4 km resolution pixel including Gothic using the Climate Engine web interface. The date ranges for deficit are a bit different from the one I sent you so that we could include 2026 on the graph.

Water deficit is computed as (Precipitation - Potential Evapotranspiration), with PET computed with the Hargreaves method.

```{r load-data}
deficit <- read.csv("./data/raw/ClimateEngine_Apr15_Jun15_Deficit_1979_2026.csv")
precip <- read.csv("./data/raw/ClimateEngine_Nov1_Apr15_Precip_1979_2026.csv")
colnames(deficit) <- c("Year","SummerDeficit")
colnames(precip) <- c("Year","WinterPrecip")
```

# Approach

I think I need to:
- Load data
- Combine the two datasets by year using a left join.
- Make a graph of Winter Precip vs Summer Deficit, with the years labeled.
- Do a T test to see if the means of the two time periods are significantly different.

# Analysis

Step 1: Combine the two datasets by year using a left join.

```{r reshape-data}

combined <- left_join(deficit,precip,by="Year")

mean_ppt_1979_2017 <- mean(combined$WinterPrecip[1:39],na.rm=TRUE)
mean_deficit_1979_2017 <- mean(combined$SummerDeficit[1:39],na.rm=TRUE)
means_df <- data.frame(label="Mean 1979-2010",
                       WinterPrecip=mean_ppt_1979_2017,
                       SummerDeficit=mean_deficit_1979_2017)

mean_ppt_2018_2026 <- mean(combined$WinterPrecip[40:48],na.rm=TRUE)
mean_deficit_2018_2026 <- mean(combined$SummerDeficit[40:48],na.rm=TRUE)
current_df <- data.frame(label="Mean 2018-2024",
                       WinterPrecip=mean_ppt_2018_2026,
                       SummerDeficit=mean_deficit_2018_2026)

hist_years <- filter(combined,Year < 2018)
cur_years <- filter(combined,Year >= 2018)
```

Step 2: Make a graph

```{r make-graph}
p1 <- ggplot(combined)+
  geom_text(aes(label=Year,x=WinterPrecip,y=SummerDeficit),hjust=0,vjust=0,data=hist_years,color="grey80",
            size=4) +
  geom_point(aes(x=WinterPrecip,y=SummerDeficit),size=3,data=means_df,color="grey80") +
  geom_point(aes(x=WinterPrecip,y=SummerDeficit),size=3,data=current_df,color="black") +
  geom_segment(aes(x=mean_ppt_1979_2017,y=mean_deficit_1979_2017,
                   xend=mean_ppt_2018_2026,yend=mean_deficit_2018_2026),
               color="red",arrow = grid::arrow(),lwd=1.5)+
  labs(title="Winter Precip vs Spring Water Deficit",
       x="Nov. 1 - May 1 Precip. (mm, water year)",
       y="Apr. 15 - Jun. 15 PPT - Et0 (mm, calendar year)") + 
  geom_text(aes(label=Year,x=WinterPrecip,y=SummerDeficit),hjust=0,vjust=0,data=cur_years,
            color="black",
            size=4) +
  theme_bw()
p1
```

Step 3: Export graph to PDF
```{r export-pdf}
pdf("./output/WinterPrecip_vs_SummerDeficit_2026_small.pdf",width=6.5,height=6)
print(p1)
dev.off()
```

Step 4: Perform a statistical test to see if the means of the two time periods are significantly different.
```{r statistical-tests}
```

# Findings

<!-- What did you learn? In your own words. -->

---

<!-- AI-use note (see handbook/04-documenting-ai-use.md):
     Briefly, how did the assistant help on this analysis? e.g.
     "AI helped debug a CRS mismatch and suggested the group_by/summarise structure;
      I wrote the plan and verified the output against the 2022 field counts." -->
